### 2020-07-17 paper.tex
remove split merge example from sec 4.2
tweak sec 4.2
remove ref to fig 1 from sec 4.1

As a concrete example, consider the probabilistic program in Figure~\ref{fig:mixture}b, which defines a Gaussian mixture model with an unknown number of components:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@gen function p(n::Int)
  k ~ poisson_plus_one(1)
  means = [
    ({(:mu, j)} ~ normal(0, 10)) for j in 1:k]
  vars = [
    ({(:var, j)} ~ inv_gamma(1, 10)) for j in 1:k]
  weights ~ dirichlet([2.0 for j in 1:k])
  for i in 1:n
    {(:x, i)} ~ mixture_of_normals(
        weights, means, vars)
  end
end
\end{lstlisting}}
\end{tabular}
\end{center}
The program \texttt{p} accepts as input an integer \texttt{n}, a number of data points. Each $\texttt{n}$ defines
a distinct distribution over dictionaries.

The first line of the program samples a number of mixture components from a Poisson prior using the address \texttt{:k}. (This line could also be written \texttt{k = \{:k\} $\sim$ poisson\_plus\_one(1)}: the \textit{address} is the symbol \texttt{:k}, and the result of the choice is assigned to a Julia variable called \texttt{k}. Because this is a common pattern, Gen provides the syntactic sugar \texttt{x $\sim$ d} as shorthand for \texttt{x = \{:x\} $\sim$ d}.) The program then samples \texttt{k} means and \texttt{k} variances from Gaussian and inverse Gamma priors, respectively. Each of these 2\texttt{k} random choices has its own address, determined by the address expression preceding it. For example, the mean for the fourth mixture component (if \texttt{k} $\geq 4$) has address \texttt{(:mu, 4)}. The mixture weights are then sampled at the address \texttt{:weights} from a Dirichlet distribution, and $\texttt{n}$ data points are sampled at addresses \texttt{(:x, 1)}, \dots, \texttt{(:x, n)}. The random choice at address $\mathtt{:}\mathtt{k}$ is discrete with $V_{\texttt{:k}} = \integers_{\ge 1}$, and the other random choices are continuous. 
Overall, the program defines a distribution over traces with the following density over traces $(A, \mathbf{x})$ with respect to $\mu_\mathcal{D}$:
\begin{align*}
p(n)((A, \mathbf{x})) =
\begin{array}{l}
p_{\mathrm{poisson}(1)}(x_{\mathtt{:}\mathtt{k}}-1) \cdot\\
\prod_{i=1}^{x_{\mathtt{:}\mathtt{k}}} p_{\mathrm{normal}(0, 10)}(x_{(\mathtt{:}\mathtt{mu}, i)}) \cdot\\
\prod_{i=1}^{x_{\mathtt{:}\mathtt{k}}} p_{\mathrm{inversegamma}(1, 10)}(x_{(\mathtt{:}\mathtt{var}, i)}) \cdot\\
p_{\mathrm{dirichlet}(2.0,\dots,2.0)}(x_{\mathtt{:}\mathtt{weights}})\cdot\\
\prod_{i=1}^n \sum_{j=1}^{x_{\mathtt{:}\mathtt{k}}} x_{\mathtt{:}\mathtt{weights}}[j] \cdot\\
\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\; p_{\mathrm{normal}(x_{\mathtt{(:mu, j)}}, x_{(\mathtt{:}\mathtt{var}, j)})}(x_{(\mathtt{:}\mathtt{x}, i)})
\end{array}
\end{align*}
when $A$ contains exactly the addresses $\mathtt{:}\mathtt{k}, \mathtt{:}\mathtt{weights}, (\mathtt{:}\mathtt{x}, i)$ for $i = 1, \dots, n$, and $(\mathtt{:}\mathtt{mu}, j)$ and $(\mathtt{:}\mathtt{var}, j)$ for $j = 1, \dots, x_{\mathtt{:}\mathtt{k}}$; otherwise, the density is 0.

@@

Recall that the involution $f$ must decompose into (i) an involution $g$ on elements $e \in E$ of a partition of the state space, and (ii) a pair of continuous differentiable bijections $h_{e}$ and $h_{g(e)} = h_{e}^{-1}$ between each pair of corresponding elements of the partition.
Each function $h_e$ is a function from the values at continuous addresses in the input trace ($z[a]$ for $a \in A_z \cap \mathcal{J}$) to the values at continuous addresses in the output trace ($z'[a]$ for $a \in A_{z'} \cap \mathcal{J}$).
We compute the Jacobian using automatic differentiation (AD);
different approaches are possible.
The Gen implementation uses forward-mode AD whereas our PyTorch implementation uses reverse-mode AD.
The procedure \textproc{run-involution} in Algorithm~\ref{alg:auto-involutive-mcmc} shows the implementation of the interpreter for the language that uses reverse-mode AD.
The interpreter executes $\mathcal{F}$ using regular Julia or Python semantics, but intercepts calls to \texttt{@write}, \texttt{@read}, and \texttt{@copy} statements, and in addition to performing the desired operation, records the set of continuous addresses that are read, written, and copied.
After $\mathcal{F}$ is finished executing, AD uses the recorded addresses to compute the Jacobian $|J h_e|(z)$; in the case of reverse-mode AD by iterating over output continuous addresses and backpropagating from each one to all of the input continous addresses, computing the Jacobian row-by-row.

%\subsubsection{Example: Involution for a Split-Merge Reversible Jump Kernel}
%
%Figure~\ref{fig:mixture}(d) shows an involution expressed in the differentiable programming language introduced in this section, and Figure~\ref{fig:mixture}(f) shows a schematic of the information flow through the involution as it is applied twice in a row.
%The involution implements a split or a merge proposal in the infinite mixture model of Figure~\ref{fig:mixture}(b), depending on the trace of the auxiliary program in Figure~\ref{fig:mixture}(c).
%If there are at least two clusters, and the value of address $\mathtt{split}$ is false, then the involution proposes to merge two clusters by running code block 2 shown in green.
%Otherwise, the involution proposes to split a cluster by running code block 1 shown in yellow.
%Note that that code block 1 (yellow) is itself an involution on the discrete addresses in the traces---each application of code block 1 to the traces causes them to flip back and forth between a split and a merge proposal.
%Also note that code blocks 2 and 3 are continuously differentiable bijections on the continuous addresses in the traces, and they are inverses of one another.

>>>>

For example, consider the probabilistic program below, which defines a Gaussian mixture model with an unknown number of components:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@gen function p(n::Int)
  k ~ poisson_plus_one(1)
  means = [
    ({(:mu, j)} ~ normal(0, 10)) for j in 1:k]
  vars = [
    ({(:var, j)} ~ inv_gamma(1, 10)) for j in 1:k]
  weights ~ dirichlet([2.0 for j in 1:k])
  for i in 1:n
    {(:x, i)} ~ mixture_of_normals(
        weights, means, vars)
  end
end
\end{lstlisting}}
\end{tabular}
\end{center}
The program \texttt{p} accepts as input an integer \texttt{n}, a number of data points. Each $\texttt{n}$ defines
a distinct distribution over dictionaries.
The first line of the program samples a number of mixture components from a Poisson prior using the address \texttt{:k}. (This line could also be written \texttt{k = \{:k\} $\sim$ poisson\_plus\_one(1)}: the \textit{address} is the symbol \texttt{:k}, and the result of the choice is assigned to a Julia variable called \texttt{k}. Because this is a common pattern, Gen provides the syntactic sugar \texttt{x $\sim$ d} as shorthand for \texttt{x = \{:x\} $\sim$ d}.) The program then samples \texttt{k} means and \texttt{k} variances from Gaussian and inverse Gamma priors, respectively. Each of these 2\texttt{k} random choices has its own address, determined by the address expression preceding it. For example, the mean for the fourth mixture component (if \texttt{k} $\geq 4$) has address \texttt{(:mu, 4)}. The mixture weights are then sampled at the address \texttt{:weights} from a Dirichlet distribution, and $\texttt{n}$ data points are sampled at addresses \texttt{(:x, 1)}, \dots, \texttt{(:x, n)}. The random choice at address $\mathtt{:}\mathtt{k}$ is discrete with $V_{\texttt{:k}} = \integers_{\ge 1}$, and the other random choices are continuous. 
Overall, the program defines a distribution over traces with the following density over traces $(A, \mathbf{x})$ with respect to $\mu_\mathcal{D}$:
\begin{align*}
p(n)((A, \mathbf{x})) =
\begin{array}{l}
p_{\mathrm{poisson}(1)}(x_{\mathtt{:}\mathtt{k}}-1) \cdot\\
\prod_{i=1}^{x_{\mathtt{:}\mathtt{k}}} p_{\mathrm{normal}(0, 10)}(x_{(\mathtt{:}\mathtt{mu}, i)}) \cdot\\
\prod_{i=1}^{x_{\mathtt{:}\mathtt{k}}} p_{\mathrm{inversegamma}(1, 10)}(x_{(\mathtt{:}\mathtt{var}, i)}) \cdot\\
p_{\mathrm{dirichlet}(2.0,\dots,2.0)}(x_{\mathtt{:}\mathtt{weights}})\cdot\\
\prod_{i=1}^n \sum_{j=1}^{x_{\mathtt{:}\mathtt{k}}} x_{\mathtt{:}\mathtt{weights}}[j] \cdot\\
\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\; p_{\mathrm{normal}(x_{\mathtt{(:mu, j)}}, x_{(\mathtt{:}\mathtt{var}, j)})}(x_{(\mathtt{:}\mathtt{x}, i)})
\end{array}
\end{align*}
when $A$ contains exactly the addresses $\mathtt{:}\mathtt{k}, \mathtt{:}\mathtt{weights}, (\mathtt{:}\mathtt{x}, i)$ for $i = 1, \dots, n$, and $(\mathtt{:}\mathtt{mu}, j)$ and $(\mathtt{:}\mathtt{var}, j)$ for $j = 1, \dots, x_{\mathtt{:}\mathtt{k}}$; otherwise, the density is 0.

@@

Recall that the involution $f$ must decompose into (i) an involution $g$ on elements $e \in E$ of a partition of the state space, and (ii) a pair of continuous differentiable bijections $h_{e}$ and $h_{g(e)} = h_{e}^{-1}$ between each pair of corresponding elements of the partition.
Each function $h_e$ is a function from the values at continuous addresses in the input trace ($z[a]$ for $a \in A_z \cap \mathcal{J}$) to the values at continuous addresses in the output trace ($z'[a]$ for $a \in A_{z'} \cap \mathcal{J}$).
We compute the Jacobian using automatic differentiation (AD);
different approaches are possible.
The Gen implementation uses forward-mode AD whereas our PyTorch implementation uses reverse-mode AD.
The procedure \textproc{run-involution} in Algorithm~\ref{alg:auto-involutive-mcmc} shows the implementation of the interpreter for the language that uses reverse-mode AD.
The interpreter executes $\mathcal{F}$ using regular Julia or Python semantics, but intercepts calls to \texttt{@write}, \texttt{@read}, and \texttt{@copy} statements, and in addition to performing the desired operation, records the set of continuous addresses that are read, written, and copied.
After $\mathcal{F}$ is finished executing, AD uses the recorded addresses to compute the Jacobian $|J h_e|(z)$; in the case of reverse-mode AD by iterating over output continuous addresses and backpropagating from each one to all of the input continous addresses, computing the Jacobian row-by-row.
Note that the $\mathtt{discrete}$ and $\mathtt{continuous}$ labels are omitted from the syntax in \textproc{run-involution} to simplify notation ($a \in \mathcal{I}$ indicates a discrete choice and otherwise a choice is continuous).
